Ecosystem size and complexity dictate riverine biodiversity

### 2021-02-22 manuscript.Rmd
Update manuscript.Rmd

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro, Greensboro, NC 27412, USA]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of
  Minnesota, St. Paul, MN 55108, USA]
- Taku Kadoya^[National Institute for Environmental Studies, Tsukuba, Ibaraki 305-8506, Japan]
- Yusuke Miyazaki^[Shiraume Gakuen College, Kodaira, Tokyo 187-8570, Japan]
output:
  html_notebook: default
abstract: The species-area relationship, a tendency in which species richness increases with increasing area, is a ubiquitous pattern in ecology[@lomolino2000] and has formed the basis of area-based conservation strategies[@desmet2004; @neigel2003]. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity)[@rodríguez-iturbe2001; @turner2015], and its role in scaling biodiversity is largely unknown. Here, we use theory and empirical analysis of extensive fish community data to show that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns[@rodríguez-iturbe2001]. Ecosystem complexity, which we define as the probability of branching per unit river distance[@terui2018; @yeakel2014], should increase watershed-scale species richness by increasing habitat heterogeneity. Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity, two factors that regulate watershed-scale species richness[@hubbell2001; @kallimanis2008]. Metacommunity simulations in branching networks predicted biodiversity patterns consistent with our hypothesis. Larger and more branched ecosystems supported greater species richness at the watershed scale, and the importance of spatial difference in species composition was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two geographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial difference in species composition. Recognizing the dual control of biodiversity should broaden viable options for spatial planning of protected areas or restoration, thereby helping conserve biodiversity from societal demands that threaten it.

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@turner2015; @rodríguez-iturbe2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodríguez-iturbe2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodríguez-iturbe2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grant2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@carrara2012; @ers2019], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@rinaldo2018; @mccluney2014]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches[@carrara2012]. However, most research has explored the consequences of branching complexity for local community structure[@ers2019; @carrara2012] or has relied solely on theoretical arguments with limited replications of branching architecture[@holt2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services[@dudgeon2006].

@@

Given the results of extensive sensitivity analysis (**Supplementary Information**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries[@moore2015; @peterson2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@poff2007; @terui2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation exists (i.e., species travel short distances). This pattern reflects significant spatial difference of species composition over the branching network. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects)[@thompson2020]. These patterns were consistent across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbell2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-18**). Under this scenario, branching may not contribute to the ecosystem's habitat heterogeneity because local environmental noise masks environmental differences between tributaries. Therefore, this landscape scenario decouples the intimate relationship between branching structure and habitat heterogeneity, thereby eliminating the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinct environmental conditions, including water temperature, water chemistry, and flow/sediment regimes[@peterson2013; @moore2015]. Instead, it may be more relevant to severely altered landscapes where human disturbance compromises the environmental distinctiveness of branches through, for example, flow regulations by dams[@poff2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: the Hokkaido island in Japan and the midwestern US (Midwest). These regions are located in comparable latitude ranges but support distinct fish communities (**Figure 1**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Methods** and **Supplementary Information**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We combined this data set with geospatial information (watershed area, branching probability, annual mean temperature, annual cumulative precipitation, the fraction of forest, and dam density) to examine potential influences of macro-scale drivers of species diversity. Using this data set, we assessed whether observed patterns are consistent across the two geographic regions by developing global and region-specific models. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity (**Methods**). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (**Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

@@

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because the watersheds in the Midwest are severely altered by agriculture (median % agricultural land use: `r round(median(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(median(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans causes significant homogenization of in-stream habitat conditions among tributaries, theory predicts a weakened effect of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial difference of species composition. In support of this interpretation, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes among tributaties. Although our analysis is correlative, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

There may be confounding factors that are not incorporated into our statistical models. However, we are confident in our statistical inference because the observed patterns were fairly consistent with our theoretical predictions that are free from any confounding effects. Further, our fitted models generally showed high R^2^ values ($\alpha$ diversity: `r round(summary(fit$alpha$model)$r.squared, 2)`; $\beta$ diversity, `r round(summary(fit$beta$model)$r.squared, 2)`; $\gamma$ diversity `r round(summary(fit$gamma$model)$r.squared, 2)`). Therefore, it is very unlikely that spurious correlations are responsible for our results.

@@

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity, ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial[@turner2015] and aquatic ecosystems[@rodríguez-iturbe2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings of the dual control of biodiversity should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. The dual dimensional approach - size and complexity - broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area). The parameter $\omega$ takes values of $0-1$ and represents the relative influence of the right tributary given the relative size of joining tributaries. With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks[@benda2004].

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro, Greensboro, NC 27412, USA]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of
  Minnesota, St. Paul, MN 55108, USA]
- Taku Kadoya^[National Institute for Environmental Studies, Tsukuba, Ibaraki 305-8506, Japan]
- Yusuke Miyazaki^[Shiraume Gakuen College, Kodaira, Tokyo 187-8570, Japan]
output: html_notebook
abstract: The species-area relationship, a tendency in which species richness increases with increasing area, is a ubiquitous pattern in ecology[@lomolino2000] and is the foundation of many conservation strategies[@desmet2004; @neigel2003]. However, many ecosystems possess a complex spatial structure that cannot be represented simply by area[@rodríguez-iturbe2001; @turner2015], and the role of  such scale-invariant complexity in scaling biodiversity relationships is largely unknown. Here, we use theory and empirical analysis of extensive fish community data to show that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns[@rodríguez-iturbe2001]. Ecosystem complexity, which we define as the probability of branching per unit river distance[@terui2018; @yeakel2014], should increase watershed-scale species richness by increasing habitat heterogeneity. Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity, two factors that regulate watershed-scale species richness[@hubbell2001; @kallimanis2008]. Metacommunity simulations in branching networks predicted biodiversity patterns consistent with our hypothesis. Larger and more branched ecosystems supported greater species richness at the watershed scale, and the importance of spatial difference in species composition was increasingly apparent when simulated species dispersal was limited. Cross-watershed comparisons across two geographic regions (Hokkaido Island in Japan and the Midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial difference in species composition. Recognizing the dual control of biodiversity should broaden viable options for spatial planning of protected areas or restoration, thereby helping conserve biodiversity from societal demands that threaten it.

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@turner2015; @rodríguez-iturbe2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodríguez-iturbe2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodríguez-iturbe2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grant2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure, and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@rinaldo2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches[@carrara2012]. However, most research has explored the consequences of branching complexity for local community structure[@ers2019; @carrara2012] or has relied solely on theoretical arguments with limited replications of branching architecture[@holt2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services[@dudgeon2006].

@@

Given the results of extensive sensitivity analysis (**Supplementary Information**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries[@moore2015; @peterson2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@poff2007; @terui2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel $\theta$), dispersal probability, and the maximum value of interspecific competition strength (see **Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation exists (i.e., species travel short distances). This pattern reflects significant spatial variation in species composition over the branching network and is likely driven by the local association of species' niche and abiotic environments (i.e., species sorting)[@thompson2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects)[@thompson2020]. These patterns were consistent across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbell2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-18**). Under this scenario, branching does not contribute to the ecosystem's habitat heterogeneity because local environmental noise masks environmental differences between tributaries. Therefore, this landscape scenario eliminates the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams with diverse geological and/or climatic backgrounds show distinct environmental conditions, including water temperature, water chemistry, and flow/sediment regimes[@peterson2013; @moore2015]. Instead, it may be more relevant to severely altered landscapes where human disturbance compromises the environmental distinctiveness of branches through, for example, flow regulations by dams[@poff2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: Hokkaido Island in Japan and the Midwestern US (Midwest). These regions are located in comparable latitude ranges but support distinct fish communities (**Figure 1**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Methods** and **Supplementary Information**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). We combined this data set with geospatial information (watershed area, branching probability, annual mean temperature, annual cumulative precipitation, the fraction of forest, and dam density) to examine potential influences of macro-scale drivers of species diversity. Using this data set, we assessed whether observed patterns are consistent across the two geographic regions by developing global and region-specific models. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity (**Methods**). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (**Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

@@

The consistent effect of branching probability on $\gamma$ diversity across the study regions is noteworthy because the watersheds in the Midwest are severely altered by agriculture (mean % agricultural land use: `r round(mean(wsd_subset[[2]]$frac_agri)*100)`% in the Midwest and `r round(mean(wsd_subset[[1]]$frac_agri)*100)`% in Hokkaido). If the intensive land use by humans causes significant homogenization of in-stream habitat conditions among tributaries, theory predicts a weakened effect of branching probability on $\gamma$ diversity. However, $\gamma$ diversity increased significantly with increasing branching probability in this highly modified landscape, suggesting that tributaries still sustain unique environmental conditions to support high spatial difference of species composition. In support of this interpretation, $\beta$ diversity increased with increasing branching probability in both regions (**Figure 4**). It is conceivable that local geological and geomorphological differences, such as slope, aspect, and soil porosity, persist in human-dominated landscapes to maintain the diversity of in-stream processes among tributaties. Although our analysis is correlative, the finding is encouraging because the branching complexity of river networks may serve as a natural defense system to human-induced environmental changes.

There may be confounding factors that are not incorporated into our statistical models. However, we are confident in our statistical inference because the observed patterns were fairly consistent with our theoretical predictions that are free from any confounding effects. Further, our fitted models showed moderate to high R^2^ values ($\alpha$ diversity: `r round(summary(fit$alpha$model)$r.squared, 2)`; $\beta$ diversity, `r round(summary(fit$beta$model)$r.squared, 2)`; $\gamma$ diversity `r round(summary(fit$gamma$model)$r.squared, 2)`). Therefore, it is very unlikely that spurious correlations are responsible for our results.

@@

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity[@iwasaki2012; @triantis2012], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial[@turner2015] and aquatic ecosystems[@rodríguez-iturbe2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings of the dual control of biodiversity should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. The dual dimensional approach - size and complexity - broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area; $A_k = 1$ at the most upstream patch). The parameter $\omega$ takes values of $0-1$ and represents the relative influence of the right tributary given the relative size of joining tributaries. With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks[@benda2004].
